\AorE{Wavelet Verstrooiing Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}

### 2024-06-22 dissertation/content/conclusion.tex
added jdp corrections

In this dissertation, we investigate the viability of wavelet transforms as alternative \ac{tf} decompositions in for detecting and classifying whale calls.

@@

From the foundation established in chapter \ref*{chap:p2} and inspired by joint-\ac{tf} scattering, we generalise the notation of using separable wavelet filters in a multi-dimensional scattering transform in chapter \ref*{chap:p3}. We show that, in a modern \ac{ml}-setting, it can be equally effective compared to other wavelet scattering filterbanks, yet provides significant advantages due to filter separability, which includes greater configuration freedoms and computational advantages.

Further work includes deploying separable wavelet scattering and the improved \ac{se} detector from chapter \ref*{chap:p2} in a \ac{pam} detection and classification system, while evaluating its performance compared to \ac{nn}-based approaches.

The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}. We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.

>>>>

In this dissertation, we investigate the viability of wavelet transforms as alternative \ac{tf} decompositions for detecting and classifying whale calls.

@@

From the foundation established in chapter \ref*{chap:p2} and inspired by joint-\ac{tf} scattering, we generalise the notion of using separable wavelet filters in a multi-dimensional scattering transform in chapter \ref*{chap:p3}. We show that, in a modern \ac{ml}-setting, it can be equally effective compared to other wavelet scattering filterbanks, yet provides significant advantages due to filter separability, which includes greater configuration freedoms and computational advantages.

Further work includes deploying separable wavelet scattering and the improved \ac{se} detec\-tor from chapter \ref*{chap:p2} in a \ac{pam} detection and classification system, while evaluating its perfor\-mance compared to \ac{nn}-based approaches.

The papers presented in this dissertation introduces wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}, while also generalising an entirely new wavelet scattering transform.

### 2024-06-22 dissertation/content/introduction.tex
added jdp corrections

At the core of many audio-based systems lies a \ac{tf}-decomposition, typically \acp{mfcc} or the \ac{stft}. This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems. At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end. Wavelet scattering features can be further improved with deep learning across filter channels \cite{nnphasecollapse}, still making it useful in a deep learning context. The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.

@@

In summary, some of the many challenges posed by \ac{pam} systems are:
\begin{enumerate}
    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in ML models if not trained correctly.
    \item Constantly evolving noise landscape, with the interference of other unpredictable bioacoustical sounds which we are often not interested in classifying.
    \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
    \item Poor \ac{snr}, often less than 0 dB.
    \item Incomplete or mislabeled training data, due to poor \ac{snr} and human error.
    \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
    \item Non-white, non-stationary background noise, which can make classification and signal detection non-trivial tasks.
    \item Poorly understood species behaviour in terms of their bioacoustics and movements.
\end{enumerate}

All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data. Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.

@@

Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter. Contextual chapters serve to unify the content of the published papers and show the natural progression of analysing the function and purpose \ac{tf} decompositions. All relevant literature is therefore gradually introduced, since each paper conceptually expanding on the previous paper's ideas and contributions.

@@

\section{Journal Papers and Contributions}

@@

>>>>

At the core of many audio-based systems lies a \ac{tf}-decomposition, typically \acp{mfcc} or the \ac{stft}. This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems. At the end of this investigative journey, we arrive at an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end. Wavelet scattering features can be further improved with deep learning across filter channels \cite{nnphasecollapse}, still making it useful in a deep learning context. The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.

@@

In summary, some of the many challenges posed by \ac{pam} systems are:
\begin{enumerate}
    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in ML models if not trained correctly.
    \item Constantly evolving noise landscape, with the interference of other unpredictable bioacoustical sounds which we are often not interested in classifying.
    \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
    \item Poor \ac{snr}, often less than 0 dB.
    \item Incomplete or mislabelled training data, due to poor \ac{snr} and human error.
    \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
    \item Non-white, non-stationary background noise, which can make classification and signal detection non-trivial tasks.
    \item Poorly understood species behaviour in terms of their bioacoustics and movements.
\end{enumerate}

All the above factors arguably discourages the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data. Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.

@@

Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter. Contextual chapters serve to unify the content of the published papers and show the natural progression of analysing the function and purpose \ac{tf} decompositions. All relevant literature is therefore gradually introduced, seeing that each paper conceptually expands on the previous paper's ideas and contributions.

@@

\section{Journal Papers}

@@

\clearpage
\section{Contributions}
This section serves as a summary of the contributions made by each paper. Contributions are ordered by most to least significant.

\begin{enumerate}
    \item \textbf{Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Trans\-form}:
    \begin{enumerate}
        \item Consider the \ac{cwt} as an alternative \ac{tf} decomposition for \ac{se} calculation.
        \item Introduce a method to scale threshold output into an interpretable and para\-meter agnostic pseudo-probability.
        \item Introduce median filtering to improve \ac{se} detectors.
    \end{enumerate}
    \item \textbf{Detecting and Classifying Whale Calls with Wavelet Scattering and Spec\-tral Entropy}:
    \begin{enumerate}
        \item Introduce \ac{ws} to the field of \ac{pam} for classification and detection tasks.
        \item Use \ac{ws} in combination with a \ac{se} detector, thereby demonstrating that classi\-fication and detection systems can coexist effectively.
        \item Improve on the previous paper's threshold scaling method by providing true proba\-bilistic output from the \ac{se} detection step.
        \item Evaluate the proposed system using practical data, providing critical discus\-sions on the reliability of typical \ac{pam} datasets. 
    \end{enumerate}
    \item \textbf{Separable Wavelet Scattering}: 
    \begin{enumerate}
        \item Introduce an entirely new family of \ac{ws} transforms, which generalises one-dimensional \ac{ws} to an arbitrary number of dimensions by using separable filters.
        \item Demonstrate \ac{sota} results on some of the MedMNIST3D \cite{medmnist} datasets, which are a collection of 3D medical datasets to benchmark \ac{ml} models, in spirit of MNIST \cite{mnist}.
    \end{enumerate}.
\end{enumerate}

### 2024-06-22 dissertation/content/nomenclature.tex
added jdp corrections

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $f(x)$ & a continuous function of $x$ \\
    $f(x, y)$ & a continuous multi-variable function of $x$ and $y$ \\
    $f[n]$ & a discrete-time (sampled) function indexed by $n$ \\
    $f[n, m]$ & a multi-variable discrete-time function indexed by $n$ and $m$ \\
    $\mathcal{F} \{f\}$ & the Fourier transform operator acting on $f$\\
    $\hat{f}$ & the Fourier transform of $f$ or an estimate of $f$, obvious from context\\
    $x^*$ & the complex conjugate of $x$\\
    $p(x)$ & a probability density \\
    $p(x|y)$ & a conditional probability density \\
    $\{a, b, ...\}$ & a set of numbers\\
    $A \setminus B$ & set minus: the set $A$ excluding the elements of $B$\\
    $A \times B$ & Cartesian product of sets\\
    $| x |$ & magnitude of a number (modulus), or size of a set\\
    $\mathbb{R}$ & the set of real numbers\\
    $\mathbb{C}$ & the set of complex numbers\\
    $\mathbb{N^+}$ & the set of natural numbers (excluding 0)\\
    $\mathbb{Z}$ & the set of integers\\
    $x \sim p$ & $x$ is drawn from the probability density $p$\\
    $\frac{d}{dx} f$ & derivative with respect to $x$ of a function $f$\\
    $\mathcal{U}_j x$ & scalogram of the $j$'th level of $x$\\
    $\mathcal{S}_j x$ & scattering coefficients of the $j$'th level of $x$\\
    $x * y$ & convolution of two functions\\
    $x \otimes y$ & circular convolution of two discrete functions\\
    $(x)_{\downarrow r}$ & a discrete function subsampled by a factor $r$\\
    $\mathbf{x}$ & a vector\\
\end{tabular}

>>>>

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $f(x)$ & a continuous function of $x$ \\
    $f(x, y)$ & a continuous multivariable function of $x$ and $y$ \\
    $f[n]$ & a discrete-time (sampled) function indexed by $n$ \\
    $f[n, m]$ & a multi-variable discrete-time function indexed by $n$ and $m$ \\
    $\mathcal{F} \{f\}$ & the Fourier transform operator acting on $f$\\
    $\hat{f}$ & the Fourier transform of $f$ or an estimate of $f$, derived from context\\
    $x^*$ & the complex conjugate of $x$\\
    $p(x)$ & a probability density \\
    $p(x|y)$ & a conditional probability density \\
    $\{a, b, ...\}$ & a set of numbers\\
    $A \setminus B$ & set minus: the set $A$ excluding the elements of $B$\\
    $A \times B$ & Cartesian product of sets\\
    $| x |$ & magnitude of a number (modulus), or size of a set\\
    $\mathbb{R}$ & the set of real numbers\\
    $\mathbb{C}$ & the set of complex numbers\\
    $\mathbb{N^+}$ & the set of natural numbers (excluding 0)\\
    $\mathbb{Z}$ & the set of integers\\
    $x \sim p$ & $x$ is drawn from the probability density $p$\\
    $\frac{d}{dx} f$ & derivative with respect to $x$ of a function $f$\\
    $\mathcal{U}_j x$ & scalogram of the $j$'th level of $x$\\
    $\mathcal{S}_j x$ & scattering coefficients of the $j$'th level of $x$\\
    $x * y$ & convolution of two functions\\
    $x \otimes y$ & circular convolution of two discrete functions\\
    $(x)_{\downarrow r}$ & a discrete function subsampled by a factor $r$\\
    $\mathbf{x}$ & a vector\\
\end{tabular}

### 2024-06-22 dissertation/content/p1_intro.tex
added jdp corrections

Derived from the conventional Fourier transform given by equation (\ref*{eqn:fouriertransform}), a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ of length $N$ samples has a \ac{dtft}
\begin{equation}
    \label{eqn:dtft}
    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
\end{equation}
for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$. The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$ \citep{dspbook}.

@@

The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast alorithm with $N \log N$ computational complexity. In particular, to circularly convolve signals $x$ and $y$, a fast computation is implemented as
\begin{equation}
    \label{eqn:fftconv}
    x \otimes y = \text{FFT}^{-1}\{\hat{x} \cdot \hat{y}\}.
\end{equation}

@@

Fourier transforms as decribed above are invertible operators, since they are orthogonal decompositions. However, for \ac{tf} analyses, invertibility is not necessarily a requirement of the application. As such, this disseration does not consider invertibility for the remainder of the content.

@@

Formally, given a signal with a fundamental frequency most accurately captured in the frequency bin $k_0$, any change in frequency by a factor $a$ will move the fundamental frequency content to the bin $a k_0$. This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of its own harmonics. As such, a small frequency shift of a lower frequency signal with ample harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

An exponential filterbank design has increased stability to small frequency fluctuations, which can make it more useful for feature-extraction methods in a \ac{ml} pipeline, since feature vectors are more stable against frequency deformations. The exponential filterbank is, in fact, the way wavelet filters are typically constructed. Additionally, the \ac{mfcc} filterbank attributes much of its sucess (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.

@@

To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially-spaced frequency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet \cite{waveletbook}. Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.

@@

The multiplicate scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.

@@

Chapter \ref{chap:p1} discusses the applications of such a filterbank in terms of the \ac{cwt}, obtaining a scalogram. Note that traditional \ac{cwt} notation typically uses the scale variable $s = \frac{1}{\lambda}$, where $s$ is made continuous. Discretisations of the \ac{cwt} approximate the transform by limiting $s$ to the set $s_k \in \frac{1}{\lambda_k}, \ \lambda_k \in \Lambda$. Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, while also using other techniques to improve traditional methods.

>>>>

Derived from the conventional Fourier transform given by equation (\ref*{eqn:fouriertransform}), a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ of length $N$ samples has a \ac{dtft}
\begin{equation}
    \label{eqn:dtft}
    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x[n] e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
\end{equation}
for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$. The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$ \citep{dspbook}.

@@

The \ac{dtft} is commonly used to quickly compute convolutions for which the \ac{fft} is a fast algorithm with $N \log N$ computational complexity. In particular, to circularly convolve signals $x$ and $y$, a fast computation is implemented as
\begin{equation}
    \label{eqn:fftconv}
    x \otimes y = \text{FFT}^{-1}\{\hat{x} \cdot \hat{y}\}.
\end{equation}

@@

Fourier transforms as described above are invertible operators, since they are orthogonal decompositions. However, for \ac{tf} analyses, invertibility is not necessarily a requirement of the application. As such, this dissertation does not consider invertibility for the remainder of the content.

@@

Formally, given a signal with a fundamental frequency most accurately captured in the frequency bin $k_0$, any change in frequency by a factor $a$ will move the fundamental frequency content to the bin $a k_0$. This has dire consequences for lower frequency signals - the frequency resolution of lower frequency signals is significantly poorer compared to the frequency resolution of that signal's harmonics. A small frequency shift of a low frequency signal with harmonic content which cause large index shifts for its harmonics, i.e., the $b$'th harmonic will be shifted to index $a b k_0$.

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to as frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}. The \ac{stft} of a signal with harmonic content is much more sensitive to such deformations across its entire structure.

An exponential filterbank design has increased stability to small frequency fluctuations, which can make it more useful for feature-extraction methods in a \ac{ml} pipeline, since feature vectors are more stable against frequency deformations. The exponential filterbank is, in fact, the way wavelet filters are typically constructed. Additionally, the \ac{mfcc} filterbank attributes much of its success (compared to raw \ac{stft} magnitude coefficients) due to its exponentially placed filters.

@@

To construct a wavelet filterbank with $\upsilon_k \in \Upsilon$ which has the exponentially spaced fre\-quency properties discussed in section \ref{sec:p1i:tfa}, we use a ``template'' wavelet $\uppsi(t)$, referred to as the mother wavelet \cite{waveletbook}. Conventionally, wavelet filters are defined as \ac{ct} filters and then discretised in software implementations.

@@

The multiplicative scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.

@@

Chapter \ref{chap:p1} discusses the applications of such a filterbank in terms of the \ac{cwt}, obtaining a scalogram. Note that traditional \ac{cwt} notation typically uses the scale variable $s = \frac{1}{\lambda}$, where $s$ is made continuous. Discretisations of the \ac{cwt} approximate the transform by limiting $s$ to the set $s_k \in \frac{1}{\lambda_k}, \ \lambda_k \in \Lambda$. Chapter \ref{chap:p1} specifically considers the application of the \ac{cwt} for signal detectors, alongside additional detector improvements.

### 2024-06-22 dissertation/content/p2_intro.tex
added jdp corrections

\ac{tf} decompositions are often used to extract features from signals. In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal. Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often present instead. In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
\begin{equation}
    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
\end{equation}

@@

The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
\begin{equation}
    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
\end{equation}
The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$. If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$. This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications \citep{cnninvariance}. In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.

@@

Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications. For example, many applications ``blur'' (low-pass) the TF decomposition \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well). The blurring operation can improve the performance of signal detectors and classifiers. However, as a cost, time localisation and/or frequency localisation is reduced.

>>>>

\ac{tf} decompositions are often used to extract features from signals. In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal. Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necessarily require a time variable: spatial variables are often present instead. In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
\begin{equation}
    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
\end{equation}

@@

The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
\begin{equation}
    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
\end{equation}
The filters $\Phi_i$ are not necessarily multi-dimensional, and may only operate along a single dimension of $X$. If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$. This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications \citep{cnninvariance}. In a sense, cascaded structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.

@@

Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications. For example, many applications ``blur'' (low-pass) the TF decom\-position \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjacent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well). The blurring operation can improve the performance of signal detectors and classifiers. However, as a cost, time localisation and/or frequency localisation is reduced.

### 2024-06-22 dissertation/content/p3_intro.tex
added jdp corrections

\section{Types of Scattering Transforms}
Existing scattering transforms are listed and described below for convenience:
\begin{enumerate}
    \item 1D scattering \citep{1dscattering1, ws}. This is the first-introduced scattering transform, which serves as a generalisation of MFCCs. It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations. Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
    \item 2D scattering \citep{2dscattering}. Gabor filters modified to have 0-mean are used to construct the filterbank. The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane. In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$ with a single filter per octave, but linearly along $\theta$.
    \item Harmonic scattering \citep{harmonicscattering} (3 dimensions). The spherical harmonics are used as filters which consitute an orthogonal basis in 3 dimensions. These filters have ``lobes'' that extend in the along various directions in the frequency domain, depending on the harmonic.
    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}. The filterbank is constructed similar to 2D scattering. In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$ with a single filter per octave, and linearly along $\theta$ and $\phi$.
    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram. A first level of scattering is performed on a 1D time signal to produce a 2D scalogram ($X[k, m]$). 2D scattering is then performed on subsequent levels, with $X[k, m]$ as input. The 2D wavelets are separable - a product of 1D wavelets. This allows 1D scattering on audio to have features that are also dependent on adjacent frequency bins, which is not possible with 1D scattering only. The extracted features can therefore correlate with frequency modulations while retaining the benefits from 1D. It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram. In the frequency domain with axes $(\omega_1, \omega_2)$, the 2D filters are constructed as the cross-product between filter sets that placed exponentially in both $\omega_1$ and $\omega_2$.
    This is the only literature considering separable filters.
\end{enumerate}

@@

For gradient descent, it may be beneficial to perform convolutions directly, and not in the Fourier domain. Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations \cite{separablecnn,separablecnn2} when computed directly. Separable filters also allow for the independance of each dimension, which can be particularly useful for optimisations such as downsampling. This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt} \cite{waveletbook}.

@@

The type of scattering transform used is usually highly dependent on the signal dimen\-sionality -- there is no unifiying multi-dimensional analogy of 1D scattering as is present with dyadic wavelet transforms \citep{waveletbook}. We construct this generalisation in the chapter \ref*{chap:p3} via separable filters.

>>>>

\section{Types of Scattering Transforms}
Existing scattering transforms are listed and described below for convenience:
\begin{enumerate}
    \item 1D scattering \citep{1dscattering1, ws}. This is the first-introduced scattering transform, which serves as a generalisation of MFCCs. It improves on MFCCs by recovering the information lost due to invariance filter (averaging) operations. Morlet wavelets are used, with cascaded scattering levels having its own number of filter per octave.
    \item 2D scattering \citep{2dscattering}. Gabor filters modified to have 0-mean are used to construct the filterbank. The filterbank consists of exponentially placed filters in one axis, which is then rotated by equally spaced angles to cover the frequency plane. In particular, in polar coordinates $(r, \theta)$ in the frequency domain, filters are placed exponentially along $r$ with a single filter per octave, but linearly along $\theta$.
    \item Harmonic scattering \citep{harmonicscattering} (3 dimensions). The spherical harmonics are used as filters which constitute an orthogonal basis in 3 dimensions. These filters have ``lobes'' that extend in the along various directions in the frequency domain, depending on the harmonic.
    \item 3D scattering -- extension of 2D scattering \citep{3dscattering}. The filterbank is constructed similar to 2D scattering. In spherical coordinates $(\rho, \theta, \phi)$ in the frequency domain, filters are spaced exponentially along $\rho$ with a single filter per octave, and linearly along $\theta$ and $\phi$.
    \item Joint time-frequency \citep{ws_joint_tf,jointtfscattering2} scattering -- 2D filters on 1D scattering scalogram. A first level of scattering is performed on a 1D time signal to produce a 2D scalogram ($X[k, m]$). 2D scattering is then performed on subsequent levels, with $X[k, m]$ as input. The 2D wavelets are separable - a product of 1D wavelets. This allows 1D scattering on audio to have features that are also dependent on adjacent frequency bins, which is not possible with 1D scattering only. The extracted features can therefore correlate with frequency modulations while retaining the benefits from 1D. It is conceptually similar to a \ac{cnn} operating on a \ac{stft} spectrogram. In the frequency domain with axes $(\omega_1, \omega_2)$, the 2D filters are constructed as the cross-product between filter sets that placed exponentially in both $\omega_1$ and $\omega_2$.
    This is the only literature considering separable filters.
\end{enumerate}

@@

For gradient descent, it may be beneficial to perform convolutions directly, and not in the Fourier domain. Separable filters have been shown to be effective in a \ac{cnn} settings, while also allowing for much quicker convolution operations \cite{separablecnn,separablecnn2} when computed directly. Separable filters also allow for the independence of each dimension, which can be particularly useful for optimisations such as downsampling. This type of scheme is illustrated by higher-dimensional dyadic wavelet transforms such as the \ac{dwt} \cite{waveletbook}.

@@

The type of scattering transform used is usually highly dependent on the signal dimen\-sionality -- there is no unifying multi-dimensional analogy of 1D scattering as is present with dyadic wavelet transforms \citep{waveletbook}. We construct this generalisation in the chapter \ref*{chap:p3} via separable filters.

### 2024-06-22 dissertation/content/paper1/main.tex
added jdp corrections

\articleinfo{Soft-output Signal Detection for Cetacean Vocalizations using
Spectral \break Entropy, K-means Clustering and the Continuous Wavelet Transform}{Ecological Informatics}{15 May 2024}{doi}

@@

\textbf{D.J.J. Verseld}: data collection; writing -- review and editing; supervision.

\textbf{J.A. Du Preez}: writing -- review and editing; supervision.

@@

\noindent which may then be converted to a pseudo-probability density $P[k,m]$, referred to as the spectral distribution:
\begin{equation}
\label{eqn:specdist}
	P[k,m] = \frac{S[k,m]}{\sum_{j=0}^{N-1} S[j,m]}.
\end{equation}

The spectral distribution is used to construct a time-varying entropy measure $H[m]$, based on Shannon's entropy:
\begin{equation}
\label{eqn:entropy}
	H[m] = - \sum_{k=0}^{N/2} P[k,m] \log (P[k,m]).
\end{equation}

@@

SE is calculated by substituting the CWT equivalent sinusoidal power from equation (\ref{eqn:cwtps}) into equations (\ref{eqn:specdist}) and (\ref{eqn:entropy}).

>>>>

\articleinfo{Soft-output Signal Detection for Cetacean Vocalizations using
Spectral \break Entropy, K-means Clustering and the Continuous Wavelet Transform}{Ecological Informatics}{May 2023}{https://doi.org/10.1016/j.ecoinf.2023.101990}

@@

\textbf{D.J.J. Versfeld}: data collection; writing -- review and editing; supervision.

\textbf{J.A. du Preez}: writing -- review and editing; supervision.

@@

\noindent which may then be converted to a pseudo-probability density $P[k,m]$, referred to as the spectral distribution:
\begin{equation}
\label{eqn:specdistp1}
	P[k,m] = \frac{S[k,m]}{\sum_{j=0}^{N-1} S[j,m]}.
\end{equation}

The spectral distribution is used to construct a time-varying entropy measure $H[m]$, based on Shannon's entropy:
\begin{equation}
\label{eqn:entropyp1}
	H[m] = - \sum_{k=0}^{N/2} P[k,m] \log (P[k,m]).
\end{equation}

@@

SE is calculated by substituting the CWT equivalent sinusoidal power from equation (\ref{eqn:cwtps}) into equations (\ref{eqn:specdistp1}) and (\ref{eqn:entropyp1}).

### 2024-06-22 dissertation/content/paper2/main.tex
added jdp corrections

\articleinfo{Detecting and classifying Whale Calls with Wavelet Scattering
and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{doi}

@@

\textbf{D.J.J. Verseld}: writing -- review and editing; supervision

\textbf{J.A. Du Preez}: writing -- review and editing; supervision.

>>>>

\articleinfo{Detecting and classifying Whale Calls with Wavelet Scattering
and Spectral Entropy}{the Journal of the Acoustical Society of America (JASA)}{15 May 2024}{TBD}

@@

\textbf{D.J.J. Versfeld}: writing -- review and editing; supervision

\textbf{J.A. du Preez}: writing -- review and editing; supervision.

### 2024-06-22 dissertation/content/paper3/main.tex
added jdp corrections

\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{doi}

@@

\textbf{D.J.J. Verseld}: writing -- review and editing; supervision.

\textbf{J.A. Du Preez}: writing -- review and editing; supervision.

>>>>

\articleinfo{Seperable Wavelet Scattering}{IEEE Signal Processing Letters}{15 May 2024}{TBD}

@@

\textbf{D.J.J. Versfeld}: writing -- review and editing; supervision.

\textbf{J.A. du Preez}: writing -- review and editing; supervision.
